Quantitative processing of broadband data as implemented in a scientific splitbeam echosounder

### 2022-05-05 Paper/Manuscript.tex
Lars changes added.

Active acoustics is an efficient method for remote sensing of marine ecosystems and is used to cover a wide range of spatial and temporal scales, ranging from tens of km \citep{makris_fish_2006} to small scale behaviour patterns \citep{klevjer_split-beam_2003}, and can provide information on key forage species, including fish and zooplankton, linking primary producers and top predators as well as other taxonomic groups important for the ecosystem functioning \citep{benoit-bird_ecological_2016}. As early as 1935, Oscar Sund observed the distribution of spawning cod in the Lofoten area using a single beam echosounder \citep{sund_echo_1935}. The method was further developed to map the abundance of fish, driven by the need for fisheries management \citep{Simmonds2005Fisheries}. More recently, active acoustics have been deployed on a wide range of platforms, including observatories, autonomous underwater vehicles \citep{fernandes_autonomous_2003}, uncrewed surface vehicles \citep{de_robertis_uncrewed_2021} and vessels of opportunity, observing a wide range of ecosystem processes across different spatial and temporal scales \citep{godo_marine_2014}.

@@

Broadband echosounders have been actively investigated for ecological research, based on the expectation of improved categorization of species or group of species. Using high frequency broadband acoustics (150-600 kHz) \citet{lavery_measurements_2010} showed that broadband signals helped to reduce the ambiguities in interpretation of acoustic scattering from zooplankton and oceanic microstructure. \citet{Stanton2012Resonance} used low frequency broadband pulses (1-6 kHz) that included the resonance frequency of swimbladdered fish, demonstrating both the spectral resolution and ability to classify size classes of fish within fish assemblages of broadband signals. \citet{blanluet_characterization_2019} used broadband, with ground truthing, to identify that the composition of two sound scattering layers (SSL) detected in the Bay of Biscay in springtime was able to be characterized, and also identified fine scale heterogeneity within the SSLs. While unsuccessful at discriminating between several species of fishes with large swimbladders during the Alaska pollock survey, \citet{bassett_broadband_2018} showed that broadband signals may be helpful in characterizing smaller fishes with swimbladders and euphausiids. \citet{benoit-bird_exploring_2020} was able to effectively discriminate three monospecific aggregations of species (hake, anchovy and krill) using broadband signals (45-170 kHz), while attempts to classify using multiple narrowband signals failed.

@@

An equally interesting outcome from the increased resolution and broad frequency bands is the ability to infer behavioural information on both single individuals, groups of individuals, and interaction between individuals or groups. \citet{Traykovski1998Effect} was able to extract orientation information about krill by using a combination of high frequency broadband acoustics and acoustic modelling. Fine-scale observations of single targets facilitated the study of interactions between prey and predators (capelin and cod) in the Barents sea \citep{skaret_diel_2020}.

@@

The theoretical foundation for signal processing comes from radar applications and has been documented for broadband echosounders \citep{stanton2008}. To tap the full potential of broadband echosounder data, further data processing methods are being developed \citep{lavery2017,bassett_broadband_2018}. Examples include adapting pulses to improve target separation as well as further parameters from the individual targets. Different transmission pulse forms can be envisioned as well as various methods for acoustic target classification. When translating equations into computer processing code, several choices must be made - some of these are well founded in signal processing literature whereas others are of a more practical and ad-hoc nature. The latter is typically missing in the literature, making it difficult to test the implementation of the signal processing methods in new echo sounders and post processing software. By presenting the design goals, implementation details, and recommended procedures and processing required to obtain quantitative broadband data, the authors hope to encourage and facilitate the realistic use of broadband signals in marine ecosystem acoustics.

The objective of this paper and associated code is to present a systematic and comprehensive description of the data processing steps for calibrated broadband echosounder data. The steps include pulse compression, target strength as a function of frequency [$\ts(\freqsym)$], and volume backscattering strength as a function of frequency [$\sv(\freqsym)$]. The intention is that the code will be used to validate implementations in various relevant data processing software, provide a starting point for further developing active acoustic broadband signal processing, and to serve as a learning resource.

@@

\begin{figure}
\includegraphics[width=16cm]{Fig_ek_sys}
\caption{\label{fi:ek_sys}Signal and data flow in the \ek{} system. An echosounder ping starts with the definition of a transmit signal (upper left) and ends with file storage (lower middle) and display and analysis (lower right). Note that all signals are complex-valued time-series.}
\end{figure}

@@

The controlling computer program generates a short-duration digital transmit signal (a ping), $\ytx(\samplesymt)$, where $\samplesymt$ is a sample index in the discrete time domain. Typical broadband pulses are linear upsweep pulses windowed by an envelope function. The generated signal $\ytxe(\timesym)$ is converted to an analogue electric signal and amplified by the transceiver to obtain an analogue signal $\ytxe(\timesym)$. The analogue and amplified signals are passed on to the transducer to generate the transmitted acoustic signal $\ytxa(\timesym)$ in the water, where $\timesym$ is the time for the (analog) signal. For a split-aperture echosounder system, there are typically three or four channels to allow estimation of the angle of arriving echoes, and the signal is typically transmitted with equal power across the channels.

In the example a linear sweep enveloped by a Hanning window is implemented (Fig.~\ref{fi:ytx}), where the parameters are the initial frequency, the final frequency, the pulse duration, the sampling rate, and the proportion of the signal that is tapered in each end, respectively. A tapering of 0 and .5 indicates no tapering and tapering across the whole signal, respectively.
\begin{figure}
\includegraphics[width=16cm]{Fig_ytx}
\caption{\label{fi:ytx} Linear chirp pulses from 34 kHz to 45 kHz with a slope of 0.057 (blue) and 0.5 (orange), as implemented in the EK80.}
\end{figure}

@@

To remove noise and reduce the quantity of data, the sampled signal from each channel is filtered and decimated in multiple stages, $\stagesym$, using complex bandpass filters, $\hfl(\genidxsym,\stagesym)$, and decimation factors, $\decfac(\stagesym)$. The individual filter coefficients for each filter and decimation stage are indexed by $\genidxsym$. The output signal from each channel, $\channelsym$, from each filter and decimation stage, $\stagesym$, is then given by:
%
\begin{equation}
\label{eq:yrx}
\yrx(\samplesymt,\channelsym,\stagesym) = \left( \yrx(\samplesymt,\channelsym,\stagesym-1) * \hfl(\genidxsym,\stagesym) \right)_{\downarrow \decfac(\stagesym)}, 
\stagesym = 1,\ldots,\nstages,
\end{equation}
%
where $\yrx(\samplesymt,\channelsym,0)$ is set to $\yrxorg(\samplesymt,\channelsym)$, being the signal before decimation, $*$ indicates convolution and $\nstages$ is the total number of filter stages. The output signal from the final filter and decimation stage, $\yrx(\samplesymt,\channelsym,\nstages)$, is shortened to $\yrx(\samplesymt,\channelsym)$ for convenience. For the output signal, $\yrx(\samplesymt,\channelsym)$, the decimated sampling rate, $\fsdec$, is given by:
%
\begin{equation}
\label{eq:fsdec}
\fsdec = \fs\prod_{\stagesym=1}^{\nstages} \frac{1}{\decfac(\stagesym)}.
\end{equation}
%
The characteristics of the bandpass filter and decimation factors are chosen with regard to the desired operating bandwidth, noise suppression levels, impulse response duration, and other common filter characteristics, with the aim of maintaining sufficient information in the data. In the example $N_v=2$. The frequency responses of the filters are shown in Figure~\ref{fi:fir} and the corresponding filter coefficients and decimation factors are given in the test data set.

@@

\subsection{Pulse compression}
To increase signal-to-noise ratio and resolution along the acoustic beam a matched filter may be applied to the raw data samples \citep{turin1960}. This technique is also known as pulse compression \citep{klauder1960}. One approach for a matched filter is to use a normalized version of the ideal transmit signal as the replicate signal, filtered and decimated using the same filters and decimation factors as applied in Eq. \ref{eq:yrx}. The normalized ideal transmit signal, $\ytxnorm(\samplesymt)$, is given by:
%
\begin{equation}
\label{eq:ytxnorm}
\ytxnorm(\samplesymt) = \frac{\ytx(\samplesymt)}{\textrm{max}(\ytx(\samplesymt))}\end{equation}
%
where $\textrm{max}$ is the maximum value of $\ytx(\samplesymt)$. The filtered and decimated output signal, $\ytxnorm(\samplesymt,\stagesym)$, from each filter stage, $\stagesym$, using the normalized ideal transmit signal, $\ytxnorm(\samplesymt)$, as the input signal, is given by:
%
\begin{equation}
\label{eq:FilterStagesTX}
\ytxnorm(\samplesymt,\stagesym) = \left[ \ytxnorm(\samplesymt,\stagesym-1) * \hfl(\genidxsym,\stagesym) \right]_{\downarrow \decfac(\stagesym)}, 
\stagesym = 1,\ldots,\nstages,
\end{equation}
%
where $\ytxnorm(\samplesymt,0)$ is set to $\ytxnorm(\samplesymt)$ and $\downarrow$ indicates decimation by the factor $\decfac(\stagesym)$. The output signal from the final filter and decimation stage, $\ytxnorm(\samplesymt,\nstages)$, is used as the matched filter and is denoted as $\ymf(\samplesymt)$  (Fig.~\ref{fi:y_mf_n}).
%
\begin{figure}
\includegraphics[width=16cm]{Fig_y_mf_n}
\caption{\label{fi:y_mf_n} The absolute value of the filtered and decimated output signal, $\ymf(\samplesymt)$, that is used for the pulse compression.}
\end{figure}

The auto correlation function of the matched filter signal and the effective pulse duration, defined as the pulse duration at transmit power $\ptxe$ which produces the same energy as the actual transmitted pulse, will be used in later processing steps and are defined as
\begin{equation}
\label{eq:TXAuto}
\ymfauto(\samplesymt) = \frac{\ymf(\samplesymt)*\ymf^*(-\samplesymt)}{||\ymf||^2_2}
\end{equation}
and
\begin{equation}
\label{eq:TauEff}
\teff = \frac{\sum \ptxauto(\samplesymt)}{\textrm{max}(\ptxauto(\samplesymt))\fsdec},
\end{equation}
where
\begin{equation*}
%\label{eq:TauEff}
\ptxauto(\samplesymt)  =  |\ymfauto(\samplesymt)|^2
\end{equation*}
is the square of the absolute value of the matched filter autocorrelation function, and the summation is calculated over a duration of twice the nominal pulse duration, $2\tnom$. For an ideal system, i.e., no tapering at the rising and trailing edges of the in-water signal, the effective pulse duration is the same as the transmit pulse duration.

@@

To perform pulse compression the received signal, $\yrx(\samplesymt,\channelsym)$, is convolved with a complex conjugated and time-reversed version of the matched filter signal with the matched filter signal, and here also normalized with the $l^2$-norm of the matched filter to maintain received signal power. The pulse compressed signal, $\ypc(\samplesymt,\channelsym)$, then becomes
\begin{equation}
\label{eq:PulseComp}
\ypc(\samplesymt,\channelsym) = \frac{\yrx(\samplesymt,\channelsym)*\ymf^*(-\samplesymt)}{||\ymf||^2_2},
\end{equation}
%
where $||\ymf||$ indicates the $l^2$-norm of $\ymf$, also known as the Euclidean norm. The received power samples are then used to estimate target strength and volume backscattering strength. For estimating received power samples, the mean signal, $\ypc(\samplesymt)$, over all transducer sectors, $\nchannels$, is used:
%
\begin{equation}
\label{eq:SumSig}
\ypc(\samplesymt) = \frac{1}{\nchannels} \sum_{\channelsym = 1}^{\nchannels} \ypc(\samplesymt,\channelsym).
\end{equation}
%
Compensation of echo strength for position in the acoustic beam requires an estimate of the echo arrival angle. This is obtained using the split-aperture method \citep{burdic1991}, which for broadband pules can be implemented with the angle values contained in the complex-valued $\ypc(\samplesymt)$ data, in combination with knowledge of transducer sector geometry. The principle is demonstrated with a transducer that is divided into four quadrants (Fig. \ref{fi:trd_quad}). In this example the summed signals from four halves (1+2, 2+3, 3+4, 4+1) are calculated as:
\begin{eqnarray}
\label{eq:SumHalves}
y_{\textrm{pc,fore}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,3)+\ypc(\samplesymt,4) \right),\\
y_{\textrm{pc,aft}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,1)+\ypc(\samplesymt,2) \right),\\
y_{\textrm{pc,star}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,1)+\ypc(\samplesymt,4) \right),\\
y_{\textrm{pc,port}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,2)+\ypc(\samplesymt,3) \right),
\end{eqnarray}
%
where fore, aft, star(board), and port indicate the relevant transducer halves.
\begin{figure}
\includegraphics{Fig_trd_quad}
\caption{\label{fi:trd_quad}Transducer divided into four quadrants. The labels are directions often used when a transducer is mounted on a ship.}
\end{figure}
%

@@

Forward/aft and port/starboard phase angles of target echoes are estimated by combining the transducer half signals thus: 
%
\begin{eqnarray}
\label{eq:phase1}
y_\along(\samplesymt) & = & y_{\textrm{pc,fore}}(\samplesymt) y_{\textrm{pc,aft}}^*(\samplesymt), \\
y_\athw(\samplesymt) & = & y_{\textrm{pc,star}}(\samplesymt) y_{\textrm{pc,port}}^*(\samplesymt),
\end{eqnarray}
%
where $y_\along(\samplesymt)$ is the electrical angle along the minor axis of the transducer (positive in the forward direction when ship-mounted) and $y_\athw(\samplesymt)$ the electrical angle along the major axis of the transducer (positive to starboard when ship-mounted), where complex signals are represented in the form $e^{j 2\pi \freqsym \timesym}$, where $j = \sqrt{-1}$. The physical echo arrival angles ($\along$ and $\athw$) are then given by:
%
\begin{eqnarray}
\label{eq:phase2}
\along(\samplesymt) & = & \arcsin\left( \frac{\atan\left( \Im(y_\along(\samplesymt)), \Re(y_\along(\samplesymt) \right)}{\anglefalong}\right) \\
\athw(\samplesymt) & = & \arcsin\left( \frac{\atan\left( \Im(y_\athw(\samplesymt)), \Re(y_\athw(\samplesymt) \right)}{\anglefathw}\right).
\end{eqnarray}
%
where $\anglefalong$ and $\anglefathw$ are constants that convert from phase angles to physical echo arrival angles and are derived from the transducer geometry and, $\fc$, the centre frequency of the chirp pulse \citep{ehrenberg1979}. The inverse sine is indicated by $\arcsin$,the four quadrant inverse tangent which returns values in the interval $[-\pi, \pi]$ inclusive is indicated by $\atan$, the real part of a complex number by $\Re$ and the imaginary part by $\Im$. As a mnemonic, the horizontal line in the symbol used for the forward/aft direction, $\along$, represents the pivot axis for the alongship angles and the near-vertical line in the $\athw$ symbol indicates the pivot axis for port/starboard angles.
%
\begin{figure}
\includegraphics[width=16cm]{Fig_theta_phi}
\caption{\label{fi:theta_phi}Physical angles $\along$ and $\athw$.}
\end{figure}

\section{Target strength}

@@

The power-budget equation (i.e., sonar equation) for a single target \citep[Formulation D, ][]{lunde2016} at frequency $\freqsym$ is:
%
\begin{equation}
\label{eq:TS}
\ts(\freqsym) = 10\log_{10}(\prxetf(\freqsym)) + 40\log_{10}(\range) + 2\absorp(\freqsym) \range 
- 10\log_{10}\left( \frac{\ptxe \wlen^2(\freqsym) \gain^2(\along_t,\athw_t,\freqsym)}{16\pi^2} \right),
\end{equation}
%
where $\prxetf(\freqsym)$ is the Fourier transform of the received electric power in a matched load for a signal from a single target at frequency $\freqsym$, $\range_t$ is the range to the target, $\absorp$ the acoustic absorption at frequency $f$, $\ptxe$ the transmitted electric power, $\wlen$ the acoustic wavelength, and $\gain$ the transducer gain incorporating both the on axis gain $\gain_0(\freqsym)=\gain(0,0,\freqsym)$ and the beam pattern based on the estimated target bearing $(\along_t,\athw_t)$.

The point scattering strength, $\mysp(\samplesymt)$, is estimated by applying Eq. \ref{eq:TS} to the received digitized power samples using the on-axis gain value with $f$ set to the centre frequency of the broadband pulse, $\fc$: 
\begin{equation}
\label{eq:Sp}
\mysp(\samplesymt) = 10\log_{10}(\prxe(\samplesymt)) + 40\log_{10}(\range(\samplesymt)) 
+ 2\absorp(\fc) \range(\samplesymt) - 10\log_{10}\left( \frac{\ptxe \wlen^2(\fc) \gainzero^2(\fc)}{16\pi^2} \right),
\end{equation}
%
noting that $\mysp(\samplesymt)$ is an average over frequency of all echoes from single or multiple targets received at sample $\samplesymt$.

@@

As a metallic sphere is a rather simple and ideal scatterer suitable for investigating TS of single scatterers, we use $\sv$ from a school of non-swimbladdered fish (Fig.~\ref{Fig_Sv_echogram}) collected with a 120 kHz centre frequency transducer to illustrate the processes to obtain $\sv(m)$.

@@

By selecting a set of centre samples $\timesym$, $\sv$ values can be presented as a function of range ($\samplesymt$) and frequency ($freqsym$) for each ping. The range for centre samples $\samplesymt$ could be chosen to half the window length or any other grid that the user prefer the data presented to be in. This can be useful when combining the $\sv(\freqsym)$ across a range of transducers. In our example we have simply chosen n to be on the same order of the range resolution (Fig.~\ref{Fig_Sv_m_n}).

@@

A set of equations for calculating calibrated, frequency-dependent, target strength and volume backscatter from broadband echosounder signals have been presented, with reference to the \ek{} echosounder.

@@

The data associated with this article are available on request from the authors.

>>>>

Active acoustics is an efficient method for remote sensing of marine ecosystems and is used to cover a wide range of spatial and temporal scales, ranging from tens of km \citep{makris_fish_2006} to small scale behaviour patterns \citep{klevjer_split-beam_2003}, and can provide information on key forage species, including fish and zooplankton, linking primary producers and top predators as well as other taxonomic groups important for the ecosystem functioning \citep{benoit-bird_ecological_2016}. As early as 1935, \citet{sund_echo_1935} observed the distribution of spawning cod in the Lofoten area using a single beam echosounder. The method was further developed to map the abundance of fish, driven by the need for fisheries management \citep{Simmonds2005Fisheries}. More recently, active acoustics have been deployed on a wide range of platforms, including observatories, autonomous underwater vehicles \citep{fernandes_autonomous_2003}, uncrewed surface vehicles \citep{de_robertis_uncrewed_2021} and vessels of opportunity, observing a wide range of ecosystem processes across different spatial and temporal scales \citep{godo_marine_2014}.

@@

Broadband echosounders have been actively investigated for ecological research, based on the expectation of improved categorization of species or group of species. Using high frequency broadband acoustics (150-600 kHz) \citet{lavery_measurements_2010} showed that broadband signals helped to reduce the ambiguities in interpretation of acoustic scattering from zooplankton and oceanic microstructure. \citet{Stanton2012Resonance} used low frequency broadband pulses (1-6 kHz) that included the resonance frequency of swimbladdered fish, demonstrating both the spectral resolution and ability to classify size classes of fish within fish assemblages of broadband signals. \citet{blanluet_characterization_2019} used broadband acoustics, with ground truthing, to identify that the composition of two sound scattering layers (SSL) detected in the Bay of Biscay in springtime was able to be characterized, and also identified fine scale heterogeneity within the SSLs. While unsuccessful at discriminating between several species of fishes with large swimbladders during the Alaska pollock survey, \citet{bassett_broadband_2018} showed that broadband signals may be helpful in characterizing smaller fishes with swimbladders and euphausiids. \citet{benoit-bird_exploring_2020} was able to effectively discriminate three monospecific aggregations of species (hake, anchovy and krill) using broadband signals (45-170 kHz), while attempts to classify species using multiple narrowband signals failed.

@@

An equally interesting outcome from the increased resolution and broad frequency information is the ability to infer behavioural information for single individuals, groups of individuals, and interaction between individuals or groups. \citet{Traykovski1998Effect} was able to extract orientation information for krill by using a combination of high frequency broadband acoustics and acoustic modelling. Fine-scale observations of single targets facilitated the study of interactions between prey and predators (capelin and cod) in the Barents sea \citep{skaret_diel_2020}.

@@

The theoretical foundation for signal processing comes from radar applications and has been documented for broadband echosounders \citep{stanton2008}. To tap the full potential of broadband echosounder data, further data processing methods are being developed \citep{lavery2017,bassett_broadband_2018}. Examples include adapting pulses to improve target separation and obtain additional parameters to describe the individual targets. Different transmission pulse forms can be envisioned as well as various methods for acoustic target classification. When translating equations into computer processing code, several choices must be made - some of these are well founded in signal processing literature whereas others are of a more practical and ad-hoc nature. The latter is typically missing in the literature, making it difficult to test the implementation of the signal processing methods in new echo sounders and post processing software. By presenting the design goals, implementation details, and recommended procedures and processing required to obtain quantitative broadband data, the authors hope to encourage and facilitate the realistic use of broadband signals in marine ecosystem acoustics.

The objective of this paper and associated code is to present a systematic and comprehensive description of the data processing steps for calibrated broadband echosounder data. The steps include pulse compression, target strength as a function of frequency [$\ts(\freqsym)$], and volume backscattering strength as a function of frequency [$\sv(\freqsym)$]. The intention is that the code will be used as a starting point for implementations in various relevant data processing software and for further developing active acoustic broadband signal processing, and to serve as a learning resource.

@@

\begin{figure}
\includegraphics[width=16cm]{Fig_ek_sys}
\caption{\label{fi:ek_sys}Signal and data flow in the \ek{} system. An echosounder ping starts with the definition of a transmit signal (upper left) and ends with file storage (lower middle) and display and analysis (lower right).}
\end{figure}

@@

The controlling computer program generates a short-duration digital transmit signal (a ping), $\ytx(\samplesymt)$, where $\samplesymt$ is the sample index in the discrete time domain. Typical broadband pulses are linear upsweep pulses windowed by an envelope function. The generated signal is converted to an analogue electric signal $\ytxe(\timesym)$ and amplified by the transceiver to obtain the analogue signal $\ytxe(\timesym)$, where $\timesym$ is the time for the signal. The analogue and amplified signals are passed on to the transducer to generate the transmitted acoustic signal $\ytxa(\timesym)$ in the water. For a split-aperture echosounder system, there are typically three or four channels to allow estimation of the angle of arriving echoes, and the signal is typically transmitted with equal power across the channels.

In the example a linear sweep enveloped by a Hanning window is implemented (Fig.~\ref{fi:ytx}), where the parameters are the initial frequency, the final frequency, the pulse duration, the sampling rate, and the proportion of the signal that is tapered in each end, respectively. A slope of 0 and .5 indicates no tapering and tapering across the whole signal, respectively.
\begin{figure}
\includegraphics[width=16cm]{Fig_ytx}
\caption{\label{fi:ytx} Linear chirp pulses from 34 kHz to 45 kHz with a slope of 0.057 (blue) and 0.5 (orange), as implemented in the EK80.}
\end{figure}

@@

To remove noise and reduce the quantity of data, the sampled signal from each channel is filtered and decimated in multiple stages, $\stagesym$, using complex bandpass filters, $\hfl(\genidxsym,\stagesym)$, and decimation factors, $\decfac(\stagesym)$. The individual filter coefficients for each filter and decimation stage are indexed by $\genidxsym$. The output signal from each channel, $\channelsym$, from each filter and decimation stage, $\stagesym$, is then given by:
%
\begin{equation}
\label{eq:yrx}
\yrx(\samplesymt,\channelsym,\stagesym) = \left( \yrx(\samplesymt,\channelsym,\stagesym-1) * \hfl(\genidxsym,\stagesym) \right)_{\downarrow \decfac(\stagesym)}, 
\stagesym = 1,\ldots,\nstages,
\end{equation}
%
where $\yrx(\samplesymt,\channelsym,0)$ is set to $\yrxorg(\samplesymt,\channelsym)$, being the signal before decimation, $*$ indicates convolution, $\downarrow$ indicates decimation by the factor $\decfac(\stagesym)$, and $\nstages$ is the total number of filter stages. The output signal from the final filter and decimation stage, $\yrx(\samplesymt,\channelsym,\nstages)$, is shortened to $\yrx(\samplesymt,\channelsym)$ for convenience. For the output signal, $\yrx(\samplesymt,\channelsym)$, the decimated sampling rate, $\fsdec$, is given by:
%
\begin{equation}
\label{eq:fsdec}
\fsdec = \fs\prod_{\stagesym=1}^{\nstages} \frac{1}{\decfac(\stagesym)}.
\end{equation}
%
The characteristics of the bandpass filter and decimation factors are chosen with regard to the desired operating bandwidth, noise suppression levels, impulse response duration, and other common filter characteristics, with the aim of maintaining sufficient information in the data. In the example $N_v=2$. The frequency responses of the filters are shown in Figure~\ref{fi:fir} and the corresponding filter coefficients and decimation factors are given in the test data set.

@@

\subsection{Pulse compression}
To increase signal-to-noise ratio and resolution along the acoustic beam a matched filter may be applied to the raw data samples \citep{turin1960}. This technique is also known as pulse compression \citep{klauder1960}. One approach for a matched filter is to use a normalized version of the ideal transmit signal as the replicate signal, filtered and decimated using the same filters and decimation factors as applied in Eq. \ref{eq:yrx}. The normalized ideal transmit signal, $\ytxnorm(\samplesymt)$, is given by:
%
\begin{equation}
\label{eq:ytxnorm}
\ytxnorm(\samplesymt) = \frac{\ytx(\samplesymt)}{\textrm{max}(\ytx(\samplesymt))}\end{equation}
%
where $\textrm{max}$ is the maximum value of $\ytx(\samplesymt)$. The filtered and decimated output signal, $\ytxnorm(\samplesymt,\stagesym)$, from each filter stage, $\stagesym$, using the normalized ideal transmit signal, $\ytxnorm(\samplesymt)$, as the input signal, is given by:
%
\begin{equation}
\label{eq:FilterStagesTX}
\ytxnorm(\samplesymt,\stagesym) = \left[ \ytxnorm(\samplesymt,\stagesym-1) * \hfl(\genidxsym,\stagesym) \right]_{\downarrow \decfac(\stagesym)}, 
\stagesym = 1,\ldots,\nstages,
\end{equation}
%
where $\ytxnorm(\samplesymt,0)$ is set to $\ytxnorm(\samplesymt)$. The output signal from the final filter and decimation stage, $\ytxnorm(\samplesymt,\nstages)$, is used as the matched filter and is denoted as $\ymf(\samplesymt)$  (Fig.~\ref{fi:y_mf_n}).
%
\begin{figure}
\includegraphics[width=16cm]{Fig_y_mf_n}
\caption{\label{fi:y_mf_n} The absolute value of the filtered and decimated output signal, $\ymf(\samplesymt)$, that is used for the pulse compression.}
\end{figure}

The auto correlation function of the matched filter signal and the effective pulse duration, defined as the pulse duration at transmit power $\ptxe$ which produces the same energy as the actual transmitted pulse, will be used in later processing steps and are defined as
\begin{equation}
\label{eq:TXAuto}
\ymfauto(\samplesymt) = \frac{\ymf(\samplesymt)*\ymf^*(-\samplesymt)}{||\ymf||^2_2}
\end{equation}
and
\begin{equation}
\label{eq:TauEff}
\teff = \frac{\sum \ptxauto(\samplesymt)}{\textrm{max}(\ptxauto(\samplesymt))\fsdec},
\end{equation}
where
\begin{equation*}
%\label{eq:TauEff}
\ptxauto(\samplesymt)  =  |\ymfauto(\samplesymt)|^2
\end{equation*}
is the square of the absolute value of the matched filter autocorrelation function, and the summation is calculated over a duration of twice the nominal pulse duration, $2\tnom$.

@@

To perform pulse compression the received signal, $\yrx(\samplesymt,\channelsym)$, is convolved with a complex conjugated and time-reversed version of the matched filter signal with the matched filter signal, and here also normalized with the $l^2$-norm of the matched filter to maintain received signal power. The pulse compressed signal, $\ypc(\samplesymt,\channelsym)$, then becomes
\begin{equation}
\label{eq:PulseComp}
\ypc(\samplesymt,\channelsym) = \frac{\yrx(\samplesymt,\channelsym)*\ymf^*(-\samplesymt)}{||\ymf||^2_2},
\end{equation}
%
where $||\ymf||$ indicates the $l^2$-norm of $\ymf$, also known as the Euclidean norm. The received power samples are then used to estimate target strength and volume backscattering strength. For estimating received power samples, the average signal, $\ypc(\samplesymt)$, over all transducer sectors, $\nchannels$, is used:
%
\begin{equation}
\label{eq:SumSig}
\ypc(\samplesymt) = \frac{1}{\nchannels} \sum_{\channelsym = 1}^{\nchannels} \ypc(\samplesymt,\channelsym).
\end{equation}
%
Compensation of echo strength for position in the acoustic beam requires an estimate of the echo arrival angle. This is obtained using the split-aperture method \citep{burdic1991}, which for broadband pules can be implemented with the angle values contained in the complex-valued $\ypc(\samplesymt)$ data, in combination with knowledge of transducer sector geometry. The principle is demonstrated with a transducer that is divided into four quadrants (Fig. \ref{fi:trd_quad}). In this example the summed signals from four halves (1+2, 2+3, 3+4, 4+1) are calculated as:
\begin{eqnarray}
\label{eq:SumHalves}
y_{\textrm{pc,fore}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,3)+\ypc(\samplesymt,4) \right),\\
y_{\textrm{pc,aft}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,1)+\ypc(\samplesymt,2) \right),\\
y_{\textrm{pc,star}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,1)+\ypc(\samplesymt,4) \right),\\
y_{\textrm{pc,port}}(\samplesymt) & = & \frac{1}{2} \left( \ypc(\samplesymt,2)+\ypc(\samplesymt,3) \right),
\end{eqnarray}
%
where fore, aft, star(board), and port indicate the relevant transducer halves.
\begin{figure}
\includegraphics{Fig_trd_quad}
\caption{\label{fi:trd_quad}Transducer divided into four quadrants. The labels are directions often used when a transducer is mounted on a ship.}
\end{figure}
%

@@

Forward/aft and port/starboard phase angles of target echoes are estimated by combining the transducer half signals thus: 
%
\begin{eqnarray}
\label{eq:phase1}
y_\along(\samplesymt) & = & y_{\textrm{pc,fore}}(\samplesymt) y_{\textrm{pc,aft}}^*(\samplesymt), \\
y_\athw(\samplesymt) & = & y_{\textrm{pc,star}}(\samplesymt) y_{\textrm{pc,port}}^*(\samplesymt),
\end{eqnarray}
%
where $y_\along(\samplesymt)$ is the electrical angle along the minor axis of the transducer (positive in the forward direction when ship-mounted) and $y_\athw(\samplesymt)$ the electrical angle along the major axis of the transducer (positive to starboard when ship-mounted), where complex signals are represented in the form $e^{j 2\pi \freqsym \timesym}$, where $j = \sqrt{-1}$. The physical echo arrival angles ($\along$ and $\athw$) are then given by:
%
\begin{eqnarray}
\label{eq:phase2}
\along(\samplesymt) & = & \arcsin\left( \frac{\atan\left( \Im(y_\along(\samplesymt)), \Re(y_\along(\samplesymt) \right)}{\anglefalong}\right) \\
\athw(\samplesymt) & = & \arcsin\left( \frac{\atan\left( \Im(y_\athw(\samplesymt)), \Re(y_\athw(\samplesymt) \right)}{\anglefathw}\right),
\end{eqnarray}
%
where $\anglefalong$ and $\anglefathw$ are constants that convert from phase angles to physical echo arrival angles (Fig.~\ref{fi:theta_phi}) and are derived from the transducer geometry and, $\fc$, the centre frequency of the chirp pulse \citep{ehrenberg1979}. The inverse sine is indicated by $\arcsin$,the four quadrant inverse tangent which returns values in the interval $[-\pi, \pi]$ inclusive is indicated by $\atan$, the real part of a complex number by $\Re$ and the imaginary part by $\Im$. As a mnemonic, the horizontal line in the symbol used for the forward/aft direction, $\along$, represents the pivot axis for the alongship angles and the near-vertical line in the $\athw$ symbol indicates the pivot axis for port/starboard angles.
%
\begin{figure}
\includegraphics[width=16cm]{Fig_theta_phi}
\caption{\label{fi:theta_phi}Physical angles $\along$ and $\athw$.}
\end{figure}
\clearpage
\section{Target strength}

@@

The power-budget equation (i.e., sonar equation) for a single target \citep[Formulation D, ][]{lunde2016} at frequency $\freqsym$ is:
%
\begin{equation}
\label{eq:TS}
\ts(\freqsym) = 10\log_{10}(\prxetf(\freqsym)) + 40\log_{10}(\range) + 2\absorp(\freqsym) \range 
- 10\log_{10}\left( \frac{\ptxe \wlen^2(\freqsym) \gain^2(\along_t,\athw_t,\freqsym)}{16\pi^2} \right),
\end{equation}
%
where $\prxetf(\freqsym)$ is the Fourier transform of the received electric power in a matched load for a signal from a single target at frequency $\freqsym$, $\range_t$ is the range to the target, $\absorp$ the acoustic absorption coefficient, $\ptxe$ the transmitted electric power, $\wlen$ the acoustic wavelength, and $\gain$ the transducer gain incorporating both the on axis gain $\gain_0(\freqsym)=\gain(0,0,\freqsym)$ and the beam pattern based on the estimated target bearing $(\along_t,\athw_t)$.

The point scattering strength, $\mysp(\samplesymt)$, is estimated by applying Eq. \ref{eq:TS} to the received digitized power samples using the on-axis gain value and $f$ set to the centre frequency of the broadband pulse, $\fc$: 
\begin{equation}
\label{eq:Sp}
\mysp(\samplesymt) = 10\log_{10}(\prxe(\samplesymt)) + 40\log_{10}(\range(\samplesymt)) 
+ 2\absorp(\fc) \range(\samplesymt) - 10\log_{10}\left( \frac{\ptxe \wlen^2(\fc) \gainzero^2(\fc)}{16\pi^2} \right),
\end{equation}
%
noting that $\mysp(\samplesymt)$ is an average over frequency of all echoes from single or multiple targets received at sample $\samplesymt$.

@@

To illustrate calculation of volume backscattering strength as a function of frequency,  $\sv(f)$, we use data collected from a school of non-swimbladdered fish (Fig.~\ref{Fig_Sv_echogram}) collected with a 120 kHz centre frequency transducer.

@@

By selecting a set of centre samples $\timesym$, $\sv$ values can be presented as a function of range ($\samplesymt$) and frequency ($freqsym$) for each ping. The range for centre samples $\samplesymt$ could be chosen to half the window length or any other grid that the user prefer the data presented to be in. This can be useful when combining the $\sv(\freqsym)$ across a range of transducers. In our example we have simply chosen the set of centre samples as the original range samples (Fig.~\ref{Fig_Sv_m_n}).

@@

A set of equations for calculating calibrated, frequency-dependent, target strength and volume backscatter from broadband echosounder signals have been presented along with example code, with reference to the \ek{} echosounder.

@@

The code and data associated with this article are available through GitHub through \url{https://github.com/CRIMAC-WP4-Machine-learning/CRIMAC-Raw-To-Svf-TSf}. The code and data for the pre-print is tagged version 0.9. The version for the printed paper is 1.0. Further developments will have higher version numbers.
